---
title: "Final Project"
author: "Annie Nguyen, Tarini Sudhakar, Soo Jee Choi"
date: "2023-04-23"
output: html_document
---

```{r setup, include=FALSE}
##Analysing trends in Billboard Top 100
library(tidyverse)
library(ClusterR) # new for kmeans++
library(foreach)
library(mosaic)
library(lubridate)
library(fastDummies)
library(tibble)
library(ggcorrplot)
library(parallel)
library(rsample) 
library(caret)
library(tidyr)
```

## IV. Patterns 
Principal Component Analysis (PCA) is a dimensionality reduction technique that can be used to identify patterns and trends in high-dimensional datasets. In the context of hit song prediction, we can use PCA to analyze how the audio and lyric features of top billboard hits have changed across decades.

First, we gathered a large dataset of songs released over several decades, including both top billboard hits. We then extract various audio features such as tempo, loudness, and danceability, as well as lyric features such as sentiment and complexity, from the songs using the Spotify Web API and natural language processing techniques. The complete data set contains the billboard top 100 hits from 1958-2015. Using the complete data set, we plot the correlation matrix for the acoustic and lyrical analysis variables, as presented below. From the matrix, as expected we see that acousticness and energy are negatively correlated while energy and loudness are positively correlated. Num_words, num_syllables, and num_lines are highly correlated, so only num_words was kept. Additionally, the various measures of sentiment (positive, negative, neutral, compound) are also highly correlated with each other, so compound_sentiment and neutral_sentiment were dropped. 

```{r all analysis, echo=FALSE}
all_data <- read.csv("https://raw.githubusercontent.com/tarinisudhakar/winningmusic/main/analysis/lyrics.csv") # 1958-2015

#Changing data integer to numeric types
all_data$danceability <- as.numeric(all_data$danceability)
all_data$energy <- as.numeric(all_data$energy)
all_data$key <- as.numeric(all_data$key)
all_data$loudness <- as.numeric(all_data$loudness)
all_data$difficult_words <- as.numeric(all_data$difficult_words)
all_data$num_dupes <- as.numeric(all_data$num_dupes)
all_data$num_words <- as.numeric(all_data$num_words)
all_data$num_lines <- as.numeric(all_data$num_lines)
all_data$duration_ms <- as.numeric(all_data$duration_ms)
all_data$mode <- as.numeric(all_data$mode)
all_data$time_signature <- as.numeric(all_data$time_signature)

#keep analysis columns
cs <- all_data[,-(33)]
cs <- cs[,-(18:19)]
cs <- cs[,-(4:5)]
cs <- cs[,-(1:2)]
cs <- na.omit(cs) # must drop na observations
cs <- cs[cs$position<=10,] #Top 10 only
X <- cs[,-(1)] # Drop position

# Center and scale the data
X = scale(X, center=TRUE, scale=TRUE)

# Extract the centers and scales from the rescaled data (which are named attributes)
mu = attr(X,"scaled:center")
sigma = attr(X,"scaled:scale")

#Correlation matrix
corr_matrix <- cor(X)
```

```{r correlation matrix, echo=FALSE}
ggcorrplot(corr_matrix)
```

For each decade, we first center and scale our data. Centering the data makes sure that the clustering algorithm focuses on the patterns and differences in the data, rather than the absolute values of each variable. Scaling the data makes sure that the PCA and clustering algorithms treats each variable equally, regardless of its magnitude or unit of measurement. 

Next, we use PCA to reduce the dimensionality of the dataset by projecting it onto a lower-dimensional subspace while preserving as much of the original variability as possible. This allows us to visualize the relationships between the different features and identify the most important factors that contribute to a song's success. We generate 10 principal components of the data for each decade. We observe that P10 explains the majority of the data, with P10 values of 84, 84, 83, 84, and 81 for the 1960s, 1970s, 1980s, 1990s, and 2000s, respectively. Each decade is known to be defined for different genre of music. For each decade, we examine at the first three principal components of the top 10 ranking songs for the decade to examine the acoustic and lyrical features of the songs during that era. The summary and full PC tables of the PCA results for each decade can be found in the appendix.

The 1960s were known for Rock ‘N Roll, Motown, and The British Invasion. Various types of Rock ‘N Roll were popular during this period such as Psychedelic Rock, Blues-Rock, and Progressive Rock. Popular singers/groups included The Beatles, Bob Dylan, The Monkees, and The Rolling Stones. We see that all three PCs are actually quite similar. They all include loudness, energy, liveness, a large number of word duplications, and negative sentiment. These may be capturing different types of Rock music popular in this period. The lack of variation in the PC1-3 results may be attributed to that fact that there were not a variety of different genres popular during this period.

```{r 1960s analysis, echo=FALSE}
#Music by decade
billboard_60s = subset(all_data, year %in% c("1960", "1961", "1962", "1963", "1964", "1965", "1966", "1967", "1968", "1969"))

#keep analysis columns
cs_60s <- billboard_60s[,-(33)]
cs_60s <- cs_60s[,-(18:19)]
cs_60s <- cs_60s[,-(4:5)]
cs_60s <- cs_60s[,-(1:2)]
cs_60s <- na.omit(cs_60s) # must drop na observations
cs_60s <- cs_60s[cs_60s$position<=10,] #Top 10 only
X_60s <- cs_60s[,-(1)] # Drop position

# Center and scale the data
X_60s = scale(X_60s, center=TRUE, scale=TRUE)

# Extract the centers and scales from the rescaled data (which are named attributes)
mu_60s = attr(X_60s,"scaled:center")
sigma_60s = attr(X_60s,"scaled:scale")

#Num_words num_syllables num_lines are highly correlated. 
X_60s = X_60s[,-(12)]
X_60s = X_60s[,-(8)]

#Sentiment is also highly correlated with itself. Dropping compound_sentiment and neutral_sentiment. 
X_60s = X_60s[,-(3:4)]
corr_matrix_60s <- cor(X_60s)

#Trying PCA
pc_sm_60s = prcomp(X_60s, rank=10, center = TRUE, scale=TRUE)

loadings_summary_60s = pc_sm_60s$rotation %>%
  as.data.frame() %>%
  rownames_to_column('Category')

#PC1
pca1_60s=loadings_summary_60s %>%
  select(Category, PC1) %>%
  arrange(desc(PC1))

#PC2
pca2_60s=loadings_summary_60s %>%
  select(Category, PC2) %>%
  arrange(desc(PC2))

#PC3
pca3_60s=loadings_summary_60s %>%
  select(Category, PC3) %>%
  arrange(desc(PC3))

# Run k-means with 3 clusters and 25 starts
clust_60s = kmeans(X_60s, 3, nstart=25)
scores_60s = pc_sm_60s$x
Y_60s = as.data.frame(cbind(X_60s, scores_60s))
cluster_60s = ggplot(Y_60s) + geom_point(aes(x=PC1, y=PC5, fill=factor(clust_60s$cluster)),
                       size=3, col="#7f7f7f", shape=21) + theme_bw()
# Examine specific songs
cluster1_songs_60s=which(clust_60s$cluster == 1) 
cluster2_songs_60s=which(clust_60s$cluster == 2) 
cluster3_songs_60s=which(clust_60s$cluster == 3)
```

```{r 1960s results, echo=FALSE}
loadings_summary_60s[c("Category","PC1","PC2","PC3")]
```

The 1970s were known for Disco, Motown, Progressive Rock, Punk, and R&B. Disco was notably one of the biggets music trends of the decade. Popular singers/groups included Diana Ross, The Jackson 5, Elton John, and Marvin Gaye. PC1, PC2, and PC3 differ considerably from each other. PC1 is largely described by acousticness, instrumentalness, and positive sentiment, perhaps picking up the rise in R&B in this decade. PC2 is described with large number of duplicate words, negative sentiment, and acousticness, perhaps a grouping of punk music that was not large, but still present in the 1970s. And finally, PC3 is described by positive sentiment, valence, danceability, energy, and loudness. This is perhaps a clear nod to the great rise in disco during the 1970s.

```{r 1970s analysis, echo=FALSE}
#Music by decade
billboard_70s = subset(all_data, year %in% c("1970", "1971", "1972", "1973", "1974", "1975", "1976", "1977", "1978", "1979"))

#keep analysis columns
cs_70s <- billboard_70s[,-(33)]
cs_70s <- cs_70s[,-(18:19)]
cs_70s <- cs_70s[,-(4:5)]
cs_70s <- cs_70s[,-(1:2)]
cs_70s <- na.omit(cs_70s) # must drop na observations
cs_70s <- cs_70s[cs_70s$position<=10,] #Top 10 only
X_70s <- cs_70s[,-(1)] # Drop position

# Center and scale the data
X_70s = scale(X_70s, center=TRUE, scale=TRUE)

# Extract the centers and scales from the rescaled data (which are named attributes)
mu_70s = attr(X_70s,"scaled:center")
sigma_70s = attr(X_70s,"scaled:scale")

#Num_words num_syllables num_lines are highly correlated. 
X_70s = X_70s[,-(12)]
X_70s = X_70s[,-(8)]

#Sentiment is also highly correlated with itself. Dropping compound_sentiment and neutral_sentiment. 
X_70s = X_70s[,-(3:4)]
corr_matrix_70s <- cor(X_70s)

#Trying PCA
pc_sm_70s = prcomp(X_70s, rank=10, center = TRUE, scale=TRUE)

loadings_summary_70s = pc_sm_70s$rotation %>%
  as.data.frame() %>%
  rownames_to_column('Category')

#PC1
pca1_70s=loadings_summary_70s %>%
  select(Category, PC1) %>%
  arrange(desc(PC1))

#PC2
pca2_70s=loadings_summary_70s %>%
  select(Category, PC2) %>%
  arrange(desc(PC2))

#PC3
pca3_70s=loadings_summary_70s %>%
  select(Category, PC3) %>%
  arrange(desc(PC3))

# Run k-means with 3 clusters and 25 starts
clust_70s = kmeans(X_70s, 3, nstart=25)
scores_70s = pc_sm_70s$x
Y_70s = as.data.frame(cbind(X_70s, scores_70s))
cluster_70s = ggplot(Y_70s) + geom_point(aes(x=PC1, y=PC5, fill=factor(clust_70s$cluster)),
                       size=3, col="#7f7f7f", shape=21) + theme_bw()
# Examine specific songs
cluster1_songs_70s=which(clust_70s$cluster == 1) 
cluster2_songs_70s=which(clust_70s$cluster == 2) 
cluster3_songs_70s=which(clust_70s$cluster == 3)
```

```{r 1970s results, echo=FALSE}
loadings_summary_70s[c("Category","PC1","PC2","PC3")]
```

The 1980s were known for Pop, Heavy Metal, Glam rock, Hip-Hop, and Country. Popular singers/groups included Whitney Houston, Madonna, Dolly Parton, Michael Jackson, Lionel Ritchie, MC Hammer, and Bon Jovi. In the 1980's music was dramatically changed by the introduction of MTV (Music Television) where a greater importance was placed on the appearance of musicians and gimmicks became commonplace. Hip-Hop also came into the mainstream during this decade. Again, PC1, PC2, and PC3 differ considerably from each other. PC1 is described by a larger number of words, energy, loudness, liveness, and speechiness, perhaps picking up the rise in hip hop during this decade. PC2 is described by neutral sentiment a larger number of words, liveness, acousticness, and instrumentalness. This grouping is more difficult to pin point, though it may be a point to country music or a grouping of pop music during this period. And finally, PC3 is described by negative sentiment, loudness, energy, higher tempo, and liveness. This may be describing a heavy metal and/or a rock grouping during this decade. 

```{r 1980s analysis, echo=FALSE}
#Music by decade
billboard_80s = subset(all_data, year %in% c("1980", "1981", "1982", "1983", "1984", "1985", "1986", "1987", "1988", "1989"))

#keep analysis columns
cs_80s <- billboard_80s[,-(33)]
cs_80s <- cs_80s[,-(18:19)]
cs_80s <- cs_80s[,-(4:5)]
cs_80s <- cs_80s[,-(1:2)]
cs_80s <- na.omit(cs_80s) # must drop na observations
cs_80s <- cs_80s[cs_80s$position<=10,] #Top 10 only
X_80s <- cs_80s[,-(1)] # Drop position

# Center and scale the data
X_80s = scale(X_80s, center=TRUE, scale=TRUE)

# Extract the centers and scales from the rescaled data (which are named attributes)
mu_80s = attr(X_80s,"scaled:center")
sigma_80s = attr(X_80s,"scaled:scale")

#Num_words num_syllables num_lines are highly correlated. 
X_80s = X_80s[,-(12)]
X_80s = X_80s[,-(8)]

#Sentiment is also highly correlated with itself. Dropping compound_sentiment and neutral_sentiment. 
X_80s = X_80s[,-(3:4)]
corr_matrix_80s <- cor(X_80s)
X_80s = X_80s[,-(21)] #No variance in time_signature
#Trying PCA
pc_sm_80s = prcomp(X_80s, rank=10, center = TRUE, scale=TRUE)

loadings_summary_80s = pc_sm_80s$rotation %>%
  as.data.frame() %>%
  rownames_to_column('Category')

#PC1
pca1_80s=loadings_summary_80s %>%
  select(Category, PC1) %>%
  arrange(desc(PC1))

#PC2
pca2_80s=loadings_summary_80s %>%
  select(Category, PC2) %>%
  arrange(desc(PC2))

#PC3
pca3_80s=loadings_summary_80s %>%
  select(Category, PC3) %>%
  arrange(desc(PC3))

# Run k-means with 3 clusters and 25 starts
clust_80s = kmeans(X_80s, 3, nstart=25)
scores_80s = pc_sm_80s$x
Y_80s = as.data.frame(cbind(X_80s, scores_80s))
cluster_80s = ggplot(Y_80s) + geom_point(aes(x=PC1, y=PC5, fill=factor(clust_80s$cluster)),
                       size=3, col="#7f7f7f", shape=21) + theme_bw()
# Examine specific songs
cluster1_songs_80s=which(clust_80s$cluster == 1) 
cluster2_songs_80s=which(clust_80s$cluster == 2) 
cluster3_songs_80s=which(clust_80s$cluster == 3)
```

```{r 1980s results, echo=FALSE}
loadings_summary_80s[c("Category","PC1","PC2","PC3")]
```

The 1990s were known for Grunge, Alternative, Rap, Hip-Hop, Gangster Rap, Bubblegum Pop, Contemporary R&B, and Country. Popular singers/groups included Nirvana, MC Hammer, Britney Spears, Backstreet Boys, N'Sync, Red Hot Chili Peppers, 2Pac, Notorious B.I.G., and Guns & Roses. Very clearly, the 1990's was an era filled with all different genre's of music. PC1 is described by acousticness, positive sentiment, longer length, and higher flesch_index scores (that is, the song lyrics are not of high difficulty to understand). This grouping may be an description of Contemporary R&B or Country. PC2 is described by loudness, energy, valence, and positive sentiment, perhaps pointing to pop songs during this period. And PC3 is described by positiveness, liveness, acousticness, and speechiness, which would describe a grouping of the Contemporary R&B present in the 90s.

```{r 1990s analysis, echo=FALSE}
#Music by decade
billboard_90s = subset(all_data, year %in% c("1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999"))

#keep analysis columns
cs_90s <- billboard_90s[,-(33)]
cs_90s <- cs_90s[,-(18:19)]
cs_90s <- cs_90s[,-(4:5)]
cs_90s <- cs_90s[,-(1:2)]
cs_90s <- na.omit(cs_90s) # must drop na observations
cs_90s <- cs_90s[cs_90s$position<=10,] #Top 10 only
X_90s <- cs_90s[,-(1)] # Drop position

# Center and scale the data
X_90s = scale(X_90s, center=TRUE, scale=TRUE)

# Extract the centers and scales from the rescaled data (which are named attributes)
mu_90s = attr(X_90s,"scaled:center")
sigma_90s = attr(X_90s,"scaled:scale")

#Num_words num_syllables num_lines are highly correlated. 
X_90s = X_90s[,-(12)]
X_90s = X_90s[,-(8)]

#Sentiment is also highly correlated with itself. Dropping compound_sentiment and neutral_sentiment. 
X_90s = X_90s[,-(3:4)]
corr_matrix_90s <- cor(X_90s)

#Trying PCA
pc_sm_90s = prcomp(X_90s, rank=10, center = TRUE, scale=TRUE)

loadings_summary_90s = pc_sm_90s$rotation %>%
  as.data.frame() %>%
  rownames_to_column('Category')

#PC1
pca1_90s=loadings_summary_90s %>%
  select(Category, PC1) %>%
  arrange(desc(PC1))

#PC2
pca2_90s=loadings_summary_90s %>%
  select(Category, PC2) %>%
  arrange(desc(PC2))

#PC3
pca3_90s=loadings_summary_90s %>%
  select(Category, PC3) %>%
  arrange(desc(PC3))

# Run k-means with 3 clusters and 25 starts
clust_90s = kmeans(X_90s, 3, nstart=25)
scores_90s = pc_sm_90s$x
Y_90s = as.data.frame(cbind(X_90s, scores_90s))
cluster_90s = ggplot(Y_90s) + geom_point(aes(x=PC1, y=PC5, fill=factor(clust_90s$cluster)),
                       size=3, col="#7f7f7f", shape=21) + theme_bw()
# Examine specific songs
cluster1_songs_90s=which(clust_90s$cluster == 1) 
cluster2_songs_90s=which(clust_90s$cluster == 2) 
cluster3_songs_90s=which(clust_90s$cluster == 3)
```

```{r 1990s results, echo=FALSE}
loadings_summary_90s[c("Category","PC1","PC2","PC3")]
```

The 2000s were known for Pop, Hip-Hop, R&B, and Rock. Popular singers/groups included Eminem, Kanye West, Nickelback, Beyoncé, Britney Spears, The Killers, and Linkin Park. The PCs all share energy and loudness features. However, PC1 is additionally described by a larger number of words, higher tempo, and more word duplications (repitition), perhaps an indication of the large presence of hip hop during this period. PC2 is additionally described by instrumentalness, liveness, valence, and negative sentiment, perhaps an indication of a section of Rock. And finally, PC3 is additionally described by liveness, higher tempo, more difficult words, and negative sentiment. This grouping is a but more difficult to pin down, though the grouping may again be referring to Rock, but perhaps a different variety (Rock was still very popular during this period with many different genres such as post-grunge, pop punk, post-hardcore, metalcore, and indie). But very clearly, the 2000s were defined by energy and loudness in their top hits.

```{r 2000s analysis, echo=FALSE}
#Music by decade
billboard_00s = subset(all_data, year %in% c("2000", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009"))

#keep analysis columns
cs_00s <- billboard_00s[,-(33)]
cs_00s <- cs_00s[,-(18:19)]
cs_00s <- cs_00s[,-(4:5)]
cs_00s <- cs_00s[,-(1:2)]
cs_00s <- na.omit(cs_00s) # must drop na observations
cs_00s <- cs_00s[cs_00s$position<=10,] #Top 10 only
X_00s <- cs_00s[,-(1)] # Drop position

# Center and scale the data
X_00s = scale(X_00s, center=TRUE, scale=TRUE)

# Extract the centers and scales from the rescaled data (which are named attributes)
mu_00s = attr(X_00s,"scaled:center")
sigma_00s = attr(X_00s,"scaled:scale")

#Num_words num_syllables num_lines are highly correlated. 
X_00s = X_00s[,-(12)]
X_00s = X_00s[,-(8)]

#Sentiment is also highly correlated with itself. Dropping compound_sentiment and neutral_sentiment. 
X_00s = X_00s[,-(3:4)]
corr_matrix_00s <- cor(X_00s)

#Trying PCA
pc_sm_00s = prcomp(X_00s, rank=10, center = TRUE, scale=TRUE)

loadings_summary_00s = pc_sm_00s$rotation %>%
  as.data.frame() %>%
  rownames_to_column('Category')

#PC1
pca1_00s=loadings_summary_00s %>%
  select(Category, PC1) %>%
  arrange(desc(PC1))

#PC2
pca2_00s=loadings_summary_00s %>%
  select(Category, PC2) %>%
  arrange(desc(PC2))

#PC3
pca3_00s=loadings_summary_00s %>%
  select(Category, PC3) %>%
  arrange(desc(PC3))

# Run k-means with 3 clusters and 25 starts
clust_00s = kmeans(X_00s, 3, nstart=25)
scores_00s = pc_sm_00s$x
Y_00s = as.data.frame(cbind(X_00s, scores_00s))
cluster_00s = ggplot(Y_00s) + geom_point(aes(x=PC1, y=PC5, fill=factor(clust_00s$cluster)),
                       size=3, col="#7f7f7f", shape=21) + theme_bw()
# Examine specific songs
cluster1_songs_00s=which(clust_00s$cluster == 1) 
cluster2_songs_00s=which(clust_00s$cluster == 2) 
cluster3_songs_00s=which(clust_00s$cluster == 3)
```

```{r 2000s results, echo=FALSE}
loadings_summary_00s[c("Category","PC1","PC2","PC3")]
```

##  V. Conclusion
While tastes for different genre of music has changed across decades, we see that certain characteristics, such as energy and loudness, continue to be important features of chart topping hits. We also observe variations in PCA PC groupings that arise as different genres of music emerged through the decades. Overall, using PCA to analyze how the audio and lyric features of top billboard hits have changed across decades can provide valuable insights into the evolution of popular music and help inform future hit song predictions.

## Appendix 1
### Data Sources
The publicly available data set by Github user `KevinSchaich` analyzed the lyrics of Billboard's Top 100 songs from 1950-2015 using a variety of Natural Language Processing techniques. The description of the variables obtained from this data set are presented below:

* year: Release year of the song
* position: Position of Billboard's Top 100 for the given year
* title: Title of the song 
* artist: Artist of the song
* pos_sentiment: Positivity association with lyrics. Value ranges between 0-1 inclusive, with 1 being 100% positive.
* neg_sentiment: Negativity association with lyrics. Value ranges between 0-1 inclusive, with 1 being 100% negative.
* neut_sentiment: Neutrality association with lyrics. Value ranges between 0-1 inclusive, with 1 being 100% neutral.
* compound_sentiment: The sum of positive, negative, and neutral scores which is then normalized between -1 (most extreme negative) and +1 (most extreme positive).
* f_k_grade: Flesch–Kincaid grade level of the song's lyrics. The Flesch-Kincaid Grade Level is equivalent to the US grade level of education.
* flesch_index: Flesch reading ease score of the song's lyrics. The Flesch reading ease score indicates the understandability of a passage with a number that ranges from 0 to 100. Higher scores indicate that the content is easier to read and understand.
* fog_index: Gunning-Fog readability index estimates the years of formal education a person needs to understand the text on the first reading. 
* num_syllables: Number of syllables in lyrics
* difficult_words: Number of words not on the Dale–Chall "easy" word list
* num_dupes: Number of duplicate (repetitive) lines in lyrics
* num_words: Number of words in lyrics
* num_lines: Number of lines in lyrics
* genre_tags: song artist's associated genre tags

The positive, negative, neutral, and compound sentiment scores of each song's lyrics were gathered using Python's Natural Language Toolkit (NLTK) VADER model. Readability metrics for each song's lyrics were obtained using the textstat package for Python. Each song artist's associated genre tags were scraped using MusicBrainz API as well as the Musicbrainzng Python interface.

Due to data availability limitations, the data set provides approximately 80-90% coverage for all the songs on Billboard's list from 1950-2015.

We obtain audio features of each song in our billboard data set using Spotify's Web API. The audio features of each song are listed below:

* danceability: Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable.
* energy: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. 
* key: The key the track is in. Integers map to pitches using standard Pitch Class notation.
* loudness: The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Values typical range between -60 and 0 db.
* mode: Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived. Major is represented by 1 and minor is 0.
* speechiness: Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording, the closer to 1.0 the attribute value.
* acousticness: A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic.
* instrumentalness: Predicts whether a track contains no vocals. The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content.
* liveness: Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. 
* valence: A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).
* tempo: The overall estimated tempo of a track in beats per minute (BPM).
* duration_ms: The duration of the track in milliseconds.
* time_signature: An estimated overall time signature of a track. The time signature (meter) is a notational convention to specify how many beats are in each bar (or measure).


## Appendix 2
### PCA 1960s 
```{r appendix 1960s, echo=FALSE}
summary(pc_sm_60s)
loadings_summary_60s
```

### PCA 1970s 
```{r appendix 1970s, echo=FALSE}
summary(pc_sm_70s)
loadings_summary_70s
```

### PCA 1980s 
```{r appendix 1980s, echo=FALSE}
summary(pc_sm_80s)
loadings_summary_80s
```

### PCA 1990s 
```{r appendix 1990s, echo=FALSE}
summary(pc_sm_90s)
loadings_summary_90s
```

### PCA 2000s 
```{r appendix 2000s, echo=FALSE}
summary(pc_sm_00s)
loadings_summary_00s
```

